
---

# Stochastic Block Model

.large[A popular probabilistic model for network data]

.pull-left[
```{tikz sbm-model, fig.cap = "The binary SBM model", fig.ext = 'png', cache=TRUE, echo = FALSE}
\usetikzlibrary{calc,shapes,backgrounds,arrows,automata,shadows,positioning}
\definecolor{myorange}{HTML}{FAB20A}
\definecolor{myblue}{HTML}{0000FF}
\definecolor{mygreen}{HTML}{008000}

\begin{tikzpicture}
  %% UN GRAPH

  \tikzstyle{every edge}=[-,>=stealth',shorten >=1pt,auto,thin,draw]
  \tikzstyle{every state}=[draw=none,text=white,scale=0.65, font=\scriptsize, transform shape]
  \tikzstyle{every node}=[fill=myorange]
  % premier cluster
  \node[state] (A1) at (0,0.5) {1};
  \node[state] (A2) at (1,0.5) {2};
  \node[state] (A3) at (.5,1.5) {3};

  \path (A2) edge [bend left] node[fill=white,below=.1cm]
  {$\pi_{\color{myorange}{\bullet}\color{myorange}{\bullet}}$}
  (A1)
  (A1) edge [bend left] (A3)
  (A3) edge [bend left] (A2);

  \tikzstyle{every node}=[fill=myblue]
  \foreach \angle/\text in {234/6, 162/7, 90/8, 18/9, -54/10} {
    \node[fill=blue,state,xshift=5cm,yshift=3.5cm]     (\text)    at
    (\angle:1cm) {\text};
  }
  \path (7) edge (10)
  (6) edge (9);
  \foreach \from/\to in {6/7,7/8,9/10,10/6}{
    \path (\from) edge [bend left] (\to);
  }

  \path    (8)    edge     [bend    left]    node[fill=white]
  {$\pi_{\color{myblue}{\bullet}\color{myblue}{\bullet}}$}  (9) ;
  
  \tikzstyle{every node}=[fill=mygreen]
  % troisieme cluster
  \node[state] (C1) at (3,-.5) {4};
  \node[state] (C2) at (4,0) {5};

  \path (C1) edge [bend right] node[fill=white,below=.25cm]
  {$\pi_{\color{mygreen}{\bullet}\color{mygreen}{\bullet}}$}
  (C2);

  % inter cluster
  \path (A3) edge [bend right]  (7)
  (A3)    edge    [bend    left]    node[fill=white]
  {$\pi_{\color{myorange}{\bullet}\color{myblue}{\bullet}}$}
  (8)
  (C2) edge [bend right] node[fill=white,right]
  {$\pi_{\color{myblue}{\bullet}\color{mygreen}{\bullet}}$}
  (9)
  (A2) edge [bend right] node[fill=white]
  {$\pi_{\color{myorange}{\bullet}\color{mygreen}{\bullet}}$}
  (C1);
\end{tikzpicture}
```
]

.pull-right[
.content-box-purple[
Let 

- *Fixed* nodes $\{1, \dots, n \}$
- Unknown colors in $\mathcal{C}=\{\color{#fab20a}{\bullet},\color{#0000ff}{\bullet},\color{#008000}{\bullet}\}$
- $\alpha_\bullet  =  \mathbb{P}(i  \in  \bullet)$, $\bullet\in\mathcal{C}$
- $\pi_{\color{#fab20a}{\bullet}\color{#0000ff}{\bullet}} = \mathbb{P}(i \leftrightarrow j | i\in\color{#fab20a}{\bullet},j\in\color{#0000ff}{\bullet})$

In other words, 

\begin{align*}
Z_i = \mathbf{1}_{\{i \in \bullet\}}  \ & \sim^{\text{iid}} \mathcal{M}(1,\alpha), \\ 
Y_{ij} \ | \ \{i\in\color{#fab20a}{\bullet},j\in\color{#0000ff}{\bullet}\}
& \sim^{\text{ind}} \mathcal{B}(\pi_{\color{#fab20a}{\bullet}\color{#0000ff}{\bullet}})\\
\end{align*}
]
]

<small>

```{r ref-sbm, results='asis', echo=FALSE}
cat("- "); print(myBib["frank1982cluster"])
cat("- "); print(myBib["holland1983stochastic"])
```

</small>
---

# Examples of topology: .small[Community network]


```{r sample-sbm-community}
pi <- matrix(c(0.3,0.02,0.02,0.02,0.3,0.02,0.02,0.02,0.3),3,3)
communities <- igraph::sample_sbm(100, pi, c(25, 50, 25))
plot(communities, vertex.label=NA, vertex.color = rep(1:3,c(25, 50, 25)))
```

---

# Examples of topology: .small[Star network]

```{r sample-sbm-star}
pi <- matrix(c(0.05,0.3,0.3,0),2,2)
star <- igraph::sample_sbm(100, pi, c(4, 96))
plot(star, vertex.label=NA, vertex.color = rep(1:2,c(4,96)))
```


---
# Estimation in the SBM: .small[latent variable model]

.pull-left[
.pull-left[
```{tikz sbm-inference, out.width='80%', fig.ext = 'png', cache=TRUE, echo = FALSE}
\usetikzlibrary{calc,shapes,backgrounds,arrows,automata,shadows,positioning}
\begin{tikzpicture}
  %% UN GRAPH

  \tikzstyle{every edge}=[-,>=stealth',shorten >=1pt,auto,thin,draw]
  \tikzstyle{every state}=[draw=none,text=white,scale=0.65, font=\scriptsize, transform shape]
  \tikzstyle{every node}=[fill=lightgray]
  % premier cluster
  \node[state] (A1) at (0,0.5) {1};
  \node[state] (A2) at (1,0.5) {2};
  \node[state] (A3) at (.5,1.5) {3};

  \path (A2) edge [bend left] node[fill=white,below=.1cm]
  {}
  (A1)
  (A1) edge [bend left] (A3)
  (A3) edge [bend left] (A2);

  \tikzstyle{every node}=[fill=blue!80!black]
  \foreach \angle/\text in {234/6, 162/7, 90/8, 18/9, -54/10} {
    \node[fill=lightgray,state,xshift=5cm,yshift=3.5cm]     (\text)    at
    (\angle:1cm) {\text};
  }
  \path (7) edge (10)
  (6) edge (9);
  \foreach \from/\to in {6/7,7/8,9/10,10/6}{
    \path (\from) edge [bend left] (\to);
  }

  \path    (8)    edge     [bend    left]    node[fill=white]
  {}  (9) ;
  
  \tikzstyle{every node}=[fill=lightgray]
  % troisieme cluster
  \node[state] (C1) at (3,-.5) {4};
  \node[state] (C2) at (4,0) {5};

  \path (C1) edge [bend right] (C2);

  % inter cluster
  \path (A3) edge [bend right]  (7)
  (A3)    edge    [bend    left]    node[fill=white]
  {}
  (8)
  (C2) edge [bend right] node[fill=white,right]
  {}
  (9)
  (A2) edge [bend right] node[fill=white]
  {}
  (C1);
\end{tikzpicture}
``` 
]
.pull-right[
<small>
.content-box-purple[
- *Fixed* nodes $\{1, \dots, n \}$
- latent colors $\mathcal{C}=\{\color{#fab20a}{\bullet},\color{#0000ff}{\bullet},\color{#008000}{\bullet}\}$
</small>
]
]
]

.pull-right[
.content-box-red[Estimate the model parameters and the clustering:
- $\theta = \{\boldsymbol\alpha = (\alpha_\bullet), \boldsymbol\Pi = (\pi_{\color{#fab20a}{\bullet}\color{#0000ff}{\bullet}})\}$
- Colors of $i$, i.e. the $\mathbf{Z}_i$
]
]
  
.large[Marginal likelihood]

Integration over $\mathcal{Z}$ is intractable: $\mathrm{card}(Q)^n$ terms!

$$p_\theta(\mathbf{Y}_i) = \int_{\mathcal{Z}} \prod_{(i,j)} p_\theta(Y_{ij} | Z_i, Z_j ) \, p_\theta(\mathbf{Z}) \mathrm{d}\mathbf{Z}$$

.large[Maximum likelihood for incomplete data model: EM]

$$\log p_\theta(\mathbf{Y}) = \mathbb{E}_{p_\theta(\mathbf{Z}\,|\,\mathbf{Y})} [\log p_\theta(\mathbf{Y}, \mathbf{Z})] + \mathcal{H}[p_\theta(\mathbf{Z}\,|\,\mathbf{Y})], \quad \text{ with } \mathcal{H}(p) = -\mathbb{E}_p(\log(p))$$ 

.important[EM requires to evaluate (some moments of)] $p_\theta(\mathbf{Z}\,|\,\mathbf{Y})$


---

# Intractable EM: solution(s)

.large[Variants of EM, MCMC/Bayesian approaches]

<small>

```{r, ref-sbm-inference, results='asis', echo=FALSE}
cat("- "); print(myBib["Nowicki2001"])
cat("- "); print(myBib["daudin2008mixture"])
cat("- "); print(myBib["latouche2012variational"])
cat("- "); print(myBib["peixoto2014efficient"])
```

</small>

.large[Variational approach]

Find a proxy $q_\psi(\mathbf{Z}) \approx p_\theta(\mathbf{Z} | \mathbf{Y})$ picked in a convenient class of distribution $\mathcal{Q}$

$$q(\mathbf{Z})^\star  \arg\min_{q\in\mathcal{Q}} D\left(q(\mathbf{Z}), p(\mathbf{Z} | \mathbf{Y})\right).$$
Küllback-Leibler  is a popular choice .small[(error averaged wrt the approximated distribution)]

$$KL\left(q(\mathbf{Z}), p(\mathbf{Z} | \mathbf{Y})\right) = \mathbb{E}_q\left[\log \frac{q(z)}{p(z)}\right] = \int_{\mathcal{Z}} q(z) \log \frac{q(z)}{p(z)} \mathrm{d}z.$$



---
# Variational EM for SBM

## Class of distribution: multinomial

$$\mathcal{Q} = \Big\{q_\psi: \, q_\psi(\mathbf{Z}) = \prod_i q_{\psi_i}(\mathbf{Z}_i), \, q_{\psi_i}(\mathbf{Z}_i) = \mathcal{M}\left(\mathbf{Z}_i; \boldsymbol\tau_i\right), \, \psi_i = \{\boldsymbol{\tau}_i\}, \boldsymbol{\tau}_i \in  \mathbb{R}^{K} \Big\}$$

Maximize the ELBO (Evidence Lower BOund):

$$J(\theta, \psi) = \log p_\theta(\mathbf{Y}) - KL[q_\psi (\mathbf{Z}) ||  p_\theta(\mathbf{Z} | \mathbf{Y})]  = \mathbb{E}_{q} [\log p_\theta(\mathbf{Y}, \mathbf{Z})] + \mathcal{H}[q_\psi(\mathbf{Z})]$$

## Variational EM

  - Initialization: get $\mathbf{T}^0 = \{\tau_{ik}^0\}$ with Absolute Spectral Clustering

  - M step: update $\theta^h = \{ \boldsymbol\alpha^h, \boldsymbol\Pi^h\}$

$$\theta^h = \arg\max J(\theta, \psi^h) = \arg\max_{\theta} \mathbb{E}_{q_{\psi^h}} [\log p_{\theta}(\mathbf{Y}, \mathbf{Z})]$$

  - VE step: find the optimal $q_\psi$, by updating $\psi^h= (\psi^h_{i})_i = \mathbf{T}^{h} = \mathbb{E}_{q^{h}} (\mathbf{Z})$:

$$\psi^h = \arg \max J(\theta^h, \psi) = \arg\min_{\psi} KL[q_\psi(\mathbf{Z}) \,||\, p_{\theta^h}(\mathbf{Z}\,|\,\mathbf{Y})]$$


---

# Variational EM for SBM: ingredients

### Variational bound


$$J(\theta, \tau ; \mathbf{Y}) = \sum_{(i,j)} \sum_{(k,\ell)} \tau_{ik} \tau_{j\ell} \log b(Y_{ij},\pi_{k\ell }) + \sum_{i} \sum_{k} \tau _{ik} \log (\alpha_k/\tau_{ik})$$


### M-step (Analytical)


$$\alpha_k = \frac{1}{n} \sum_{i} \tau_{i k} , \quad  \pi_{k\ell } = \frac{\sum_{(i,j)} \tau_{ik}\tau_{j\ell} Y_{ij}}{\tau_{ik}\tau_{j\ell}} \qquad \left({\boldsymbol\alpha} = \mathbf{1}_n^\top\mathbf{T}, \quad {\boldsymbol\Pi} =  \frac{\mathbf{T}^\top \mathbf{Y} \mathbf{T}}{\mathbf{T}^\top  \mathbf{T}} \right)$$

### Variational E-step (fixed point)


$$\tau_{ik} \varpropto \alpha_k \prod_{(i,j)} \prod_{\ell} b(Y_{ij} ; \pi_{k\ell})^{\tau_{j\ell}}$$

### Model Selection

$$\mathrm{vICL}(K) = \mathbb{E}_{q} [\log L(\hat{\theta)}; \mathbf{Y}, \mathbf{Z}] - \frac{1}{2} \left(\frac{K(K+1)}{2} \log \frac{n(n-1)}{2} + (K-1) \log (n) \right)$$


---

# Example: French politcal blogosphere

```{r fblog-simpleSbm-analysis, cache = TRUE}
blog  <- as_adj(frenchblog2007, sparse = FALSE)
blocks <- 1:18
sbm_full <- estimateMissSBM(blog, blocks, "node")
```

---

# Convergence monitoring (ELBO)

```{r fblog-simpleSbm-analysis-plot1}
plot(sbm_full, "monitoring")
```

---

# Model Selection (vICL)

```{r fblog-simpleSbm-analysis-plot2}
plot(sbm_full)
```

---

# Parameters

```{r fblog-simpleSbm-analysis-theta}
plot(sbm_full$bestModel, "meso")
```

---

# Clustering I

```{r fblog-simpleSbm-analysis-plot3}
plot(sbm_full$bestModel, dimLabels = list(row = "blogs", col = "blogs"))
```

---

# Clustering II

```{r fblog-simpleSbm-analysis-plot4}
plot(sbm_full$bestModel, "expected", dimLabels = list(row = "blogs", col = "blogs"))
```


---

# Clustering III

```{r fblog-simpleSbm-analysis-ARI}
aricode::ARI(sbm_full$bestModel$fittedSBM$memberships, party)
aricode::NID(sbm_full$bestModel$fittedSBM$memberships, party)
```


