---
title: "missSBM"
subtitle: "Inference in Stochastic Block Models from Missing Data"
author: " P. Barbillon, J. Chiquet, T. Tabouy<br /><br /> <small>Paris-Saclay, AgroParisTech, INRAE</small> <br /> <small>Last update `r format(Sys.time(), '%d %B, %Y')`</small>"
date: "<br/>https://github.com/GrossSBM/missSBM"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    css: ["my_pres.css", default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.align='center')
library(tidyverse) # data manipulation
library(corrplot)  # plot of covariance/correlation matrices
library(ggraph)
library(igraph)
library(RColorBrewer)
library(aricode)

fancy_network <- function(mygraph, group) {

  V(mygraph)$party  <- as.character(group)
  V(mygraph)$degree <- degree(mygraph)

  angle <- 90 - 360 * 1:igraph::gorder(mygraph) / igraph::gorder(mygraph)
  V(mygraph)$hjust <- ifelse(angle < -90, 1, 0)
  V(mygraph)$angle <- ifelse(angle < -90, angle + 180, angle)
  
  p <- ggraph(mygraph, layout = 'linear', circular = TRUE) + 
    geom_edge_arc(alpha = 0.25, show.legend = FALSE) +
#    geom_node_text(aes(label = party, x = x*1.05, y=y*1.05, angle = angle, hjust = hjust), size = 3) +
    geom_node_point(aes(colour = party, size = degree), alpha = 0.5) +
    scale_colour_manual(values = c(brewer.pal(6,"Paired"),"maroon4","springgreen", "tan4","plum","brown4","burlywood",brewer.pal(8, "Dark2"), "navy")) +
    theme_graph(base_size = 20)
  
  invisible(p)
}
```

```{r, load_refs, echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE}
library(RefManageR)
BibOptions(check.entries = FALSE, 
           bib.style = "authoryear", 
           cite.style = 'alphabetic', 
           style = "markdown",
           hyperlink = FALSE, 
           dashed = FALSE)
myBib <- ReadBib("./missSBM.bib", check = FALSE)
```

# Resources

### R/C++ package

Last stable release on CRAN, development version available on GitHub.

```{r install, eval=FALSE, tidy=FALSE}
install.packages("missSBM")
remotes::install_github("GrossSBM/missSBM@development")
```

```{r loading package, message=FALSE}
library(missSBM)
packageVersion("missSBM")
```

```{r future, echo = FALSE}
options(future.fork.enable = TRUE)
future::plan("multicore", workers = parallel::detectCores())
```

### Publications

<small>

The [missSBM website](https://grosssbm.github.io/missSBM/) contains the standard package documentation and a couple of vignettes for the top-level functions.

```{r, results='hide', echo = FALSE}
RefManageR::Cite(myBib, 'Tabouy2019')
RefManageR::Cite(myBib, 'misssbm')
```

```{r, 'refs', results='asis', echo=FALSE}
cat("- "); print(myBib["Tabouy2019"])
cat("- "); print(myBib["misssbm"])
```

</small>


---

class: inverse, middle

# Outline 

  1. .large[Motivations]
  2. .large[Binary SBM and variational Inference]
  3. .large[SBM inference from observed data]
  4. .large[Illustration]

---

# Network data with missing entries

### Recommandation system: Epinion 

Who-trust-whom online social network of a general consumer review site Epinions.com. Members of the site can decide whether to ''trust'' each other. 

**Available** at http://www.trustlet.org/datasets/extended_epinions/user_rating.txt.gz

### Social networks in ethnobiology

A seed exchange network in Kenya is collected on a limited space area, where all the 155 farmers are interviewed. Farmers only provide information about other farmers with whom they have interacted.

### Ecological networks: plant-pollinator nbetwork

Interaction network between predefined sets of plants and pollonitor, by direct observation. How can trust the "0" in network data collected not to be a missing entry? 

---

# .small[Companion data set: French political Blogosphere]

Single day snapshot of almost 200 political blogs automatically extracted the 14 October 2006 and manually classified by the "Observatoire Présidentielle" project.


```{r, eval=FALSE}
data("frenchblog2007")
party <- vertex.attributes(frenchblog2007)$party
table(party) %>% kableExtra::kbl() %>% kableExtra::kable_classic()
```


```{r, echo = FALSE}
data("frenchblog2007")
frenchblog2007 <- delete_vertices(frenchblog2007, which(degree(frenchblog2007) == 0))
party <- vertex.attributes(frenchblog2007)$party
table(party) %>% kableExtra::kbl() %>% kableExtra::kable_classic()
```

---

# French blog: graph view

```{r blog-graph, cache = TRUE, echo = FALSE, fig.align='center', fig.dim=c(9,7), out.width='95%'}
print(fancy_network(frenchblog2007, party))
```

---

# French blog: matrix view

```{r blog-matrix, cache = TRUE, echo = FALSE, fig.align='center', fig.dim=c(7,7), out.width='70%'}
blog  <- as_adj(frenchblog2007, sparse = FALSE)
sbm::plotMyMatrix(blog, dimLabels = list('blog', "blog ordered per party"), clustering = list(row = party))
```

---

class: inverse, middle

# SBM: background

  - Probabilistic model for random graph
  - Latent variable model
  - Variational Inference

<!-- SBM: background -->
```{r SBM_background, child='sbm.Rmd'}
```

---

class: inverse, middle

# SBM from an observed network

  - Missing data framework for SBM
  - Modeling the observation process
  - Inference with missing dyads

<!-- MISS-SBM -->
```{r missSBM, child='missSBM.Rmd'}
```

---

class: inverse, middle

# Illustrations

  1. Numerical study of MNAR vs MAR
  2. French blogosphere
  3. PPI ER (ESR1) ego network

<!-- Illustration MNAR -->
```{r MNAR, child='simus.Rmd'}
```

<!-- Illustration blogosphere -->
```{r blog, child='fblog.Rmd'}
```

<!-- PPI-network -->
```{r blog, child='ppi-network.Rmd'}
```


```{r future off, echo=FALSE}
future::plan("sequential")
```

---
# Conclusion

## Perspectives/ongoing

- Sampling
  - study robustness (block-sampling "includes" double-standard?)
- Other models
  - degree-corrected SBM
  - (ZI)-Poisson emission law
  - Simple SBM $\to$ Bipartite SBM (aka Latent block models)
- Other algorithms
  - SGD algorithms + Pytorch framework
  - 'Exact' ICL maximization (with É. Côme)


## Advertisement

[https://computo.sfds.asso.fr](https://computo.sfds.asso.fr), a new journal promoting reproducible research

## THANK YOU

---
# References

```{r, 'print_refs', results='asis', echo=FALSE}
PrintBibliography(myBib)
```

